Efficiency

### 2018-01-08 DraftExec.tex
next

A comprehensive description of the development of the propeller can be found in \cite{carlton2007}.  Here I won't elaborate on beyond describing some essential features and characteristics with which one ought to be familiar, primarily for the purpose of comparing different propellers and their evolution into impellers of different designs.

@@

\subsubsection{Pitch}
Perhaps one of the most important characteristics of a propeller is its pitch.  The pitch represents the distance that a blade section would travel forward if it carved its way around a helix through one full rotation.  It is intuitive to think of as something akin to the angle-of-attack of the propeller blade to the water, however this is technically misleading as the movement of the water-column incoming to the blade, as well as the speed of rotation, also have a significant bearing on what the actual angle of attack of the blade ends up being.  Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unitless equivalent, as is common in the description of propellers and propulsion systems.

@@

The selection of the exact blade selection that is selected to define pitch must be specified as being at some fraction of the radius from the centre.  This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.  A thorough discussion of pitch can be found in \cite[35-37]{carlton2007}.

@@

Where $V_a$ is the speed of advance of the propeller, or the speed at which water arrives at the front-side of the propeller, and $n$ represents the rate of the propeller's rotation (revolutions per second) and $D$ is the propeller's diameter.  For fullness of understanding it is also worthwhile to note here that $V_a$ does not correspond directly to the vessel's speed of travel, though they are related. To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed \parencite[15]{MAN2017}.  Another key distinction for the consideration of jets later on, is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field. In \cite[213]{lewis1988} $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.

One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$. This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced. This represents circumstances where a propeller encounters effectively zero load.  This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.

@@

\subsection{Ducted Propellers}
A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.  Ducted propellers fall into broadly two categories, one of which is referred to frequently and interchangeably as a pump-jet.  The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating type.  The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).

A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted. (Aside: It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.  But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear.)

@@

\subsubsection{The Accelerating Duct}
Accelerating ducts are most commonly used by vessels which have to operate at high loadings at very low speed, such as tugs, barges, or trawlers.  In those circumstances, relatively small vessels need to exert considerable forces against substantial loads, at very low water-speeds.  Also, in these circumstances, the practical size of a propeller might be constrained, since operating an optimally sized propeller might increase the complexity of machiner required to deliver the power to so low a hub in the water, and it might also be difficult to arrange for the thrust to be directed substantially against the centre of drag of the vessel.

The effect of the nozzle in these circumstances actually increase the optimal efficiency of the propeller, when operating at very low water-speeds, since the pressure reduction on the inside of the nozzle induced by the movement of the water through the nozzle, actually produces a net thrust forward, since the interior face at the opening of the nozzle is necessarily angled forward. This has the effect of distributing the suction force across a larger area, which now includes the interior of the nozzle.  An equivalent way of describing the same effect is that the nozzle causes the water to be accelerated more gradually, over a longer distance, by inducing it to move more quickly as it passes through to the entrance.  As a broad generalisation, more gradual or gentle actions tend to involve lower overall losses to turbulence, and hence the efficiency of the propeller can be increased.

A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.  Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.  It is also around this vortex that blade tip cavitation tends to occur. By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation can be diminished.

However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.  In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased, and cavitation can and does still occur in ducted propellers near the blade tips \parencite{moulijn2015}.

@@

A detailed explanation of the means by which the ideal efficiency of propeller can be improved can be found in \cite[213-222]{lewis1988}, as well as other texts on marine propulsion, such as \cite{carlton2007}. A substantial work exploring the efficiencies of a range of accelerating and decelerating ducts can also be found in \cite{oosterveld1970}.  A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low waterspeeds.  At higher waterspeeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.  When the exterior of the duct experience relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.  Furthermore, in general these models do not incorporate the effect of flow separation, which is acknowledged as a possibility on both the exterior and interior of ducts, which can dramatically reduce efficiency performance (\cite[20]{oosterveld1970}, \cite[214]{lewis1988}).

@@

In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract at this point.

The addition of some new part such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers. The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.

\subsubsection{Waterjets}
The use of the word waterjet, as opposed to pumpjet, signifies that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.  The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.  However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump. Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).  There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}.

@@

\subsection{Recent Commentary}
There has been already some attention given to the probable efficiencies of pump jets at low speed. Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.  This argument relies pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.  As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.  This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pump-jet for a submarine might perform in quantitative terms, relative to a propeller.

There has also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017} and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.  Whilst the argument that the state of the art is much advanced is obviously true, it is the case with marine propulsion, as with most sciences, that new developments tend to confirm, rather than annul, many of the workings that have been previously undertaken.  To the extent that modern literature still supports the overall constraints and bounds provided by some of the foundational earlier works, a great deal of useful information can still be obtained from earlier works.  In particular, the foundational theories, often derived from well-established principals of conservation of momentum, energy, as applied by the likes of Bernoulli and Euler, are as well respected and deeply depended upon today as they were centuries earlier when they were first arrived at.

\subsection{Modern literature and results}
In addition, more modern works, including those which involve experimental validation of theoretical results also clearly confirm that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.  It should be noted however, that relatively little literature focuses on the performance of propulsive systems far far outside of their intended design speeds, which tends to be far higher than the patrol and transit speeds of diesel electric submarines.  Fewer still tend to express their work in terms of ship-speed, instead adopting the convetion of using dimensionless ratios so that work done on one system can be readily translated into different combinations loads, pressures, speeds and sizes.  However, from a few notable studies exist which provide important reference points. (Further interpretation of the what we can lean from the trends expressed in dimensionless coefficients will be discussed later.)

@@

Both of these previous studies confirm the general and widespread assumption that jets are known to be more efficient at high speeds, and propellers at lower speeds.  They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.  However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets). Pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.  Given that these factors might affect efficiency, and the overall design of the nozzle considerably, more direct comparison with fully submerged jets should be sought to make more direct comparisons.

@@

The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.  The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.  In the case of this analysis, modifications to the hullform in order to accommodate the jet (which are likely to be more substantial for a surface ship than a submarine) also have an impcat on the resistance offerd by the hull.  This also has an interaction with the different shape of the propulsors, which alters the required level of delivered (effective) power to achieve a given speed, in addition to the considerations of how efficiently the propulsor generates that effective power. The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form. If the power could be reduced by 5\% (e.g. by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hullform needed to accommodate the jet, rather than the jet itself. \parencite{giles2010}

The paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.  At 5kt, the propeller was more than twice as efficient as the waterjet.

@@

It appears from recent examples in the literature that there is no evidence of jets approaching the same efficiency as propellers at speeds as low as 10kt.  Indeed, there seems to be considerable evidence that jets of most kinds suffer substantial efficiency penalties relative to propellers in this regime, and that this is understood and accepted as an inevitable design trade-off to be made.  In the case of a submarine, which might spend nearly 100\% of its time travelling at 10kt or less, this disadvantage (as opposed to less than 35\%, as assumed for a surface ship) would become extremely substantial, which will be quantified in later sections.

Moreover, recent literature seems to confirm that efficiency of pumpjets fall off precipitously towards zero the closer that the vessel speed comes to zero.  This strongly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.  To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.  This should further limit the range of error that might be made in estimating realistic bounds of performance for an unknown system.

@@

Pumpjets are necessarily low in efficiency at very low speeds because they attempt to get the water to do multiple things.  They don't just accelerate a water colum.  They first slow it down (increasing its pressure), then add pressure to it, then accelerate it.  When the total amount of energy that the incoming water has, and the total amount of energy you want to add to the water both get very small (as they do at very low speed) there simply isn't enough energy to do all things at once to all of the water.  The energy penalty inflicted by using the additional devices becomes very large in proportion, and part of the water doesn't do all the necessary things for the jet to work.  In essence, if you try to slow down something that's already going slowly, it tends to stop.  That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, takes place.

There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in the next subsection.  They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed.  the the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation, which is what occurs when the assumptions inherent in momentum theory's analysis of an 'ideal' actuator disc start to break down.

The conditions which lead to flow separation in a pumpjet can be relatively simply explained in terms of conservation of energy, and conservation of mass. Conservation of mass requires that as a pipe or duct expands, the flow through it necessarily slows down, in order to occupy the large volume, in proportion to the new cross-sectional area.  Conservation of energy requires that it increases in pressure, and defines a very specific relationship between the pressure increase and the change in velocity, which is determined by the shape of the duct.  That relationship says that the static pressure increase is inversely proportional change in velocity **squared**, as given by Bernoulli's equation, which was discussed in earlier sections.  As a result, at high speeds, the total energy of the system is dominated by its velocity, but at very low speeds the static pressure is much more significant.

It also means that for any given change in velocity, the static pressure changes also, but doesn't change as much as the velocity change.  The consequences for this pumpjets are and inevitable, since the geometry of the duct is inevitably fixed.  At very high flow rates, the expansion of the area ahead of the impeller creates an adverse pressure gradient, as the water must slow down to fill the greater area.  This adverse pressure gradient is larger at higher speeds, the proportions aren't linear, hence the the speed is much much larger, due to the square relationship.

@@

The same phenomenon is also prone to occur near roots of the blades of the rotor.  Necessarily, the blades of the rotor are doing work on the fluid, so unless the impeller is in a region of considerable contraction, pressure will necessarily have to increase through the blades, providing an adverse pressure gradient, which again must be constant radially.  Since closer to the hub the velocity of the blads through the water is lower, the blades tend often to also be longer and deflect the flow even more in order to do the same amount of work on the fluid. With the longer surface over which the fluid must travel, as well as greater angle of deflection, cavitation is prone to occur here too.  The physics will be exactly analogous to that of the separation on the inside of the duct intake, except the pressure gradient will be provided by a different source (the rotor doing work), and the surface (the blade roots) will also be different. This will also mean that the direction of rotation of the vortices will be different, and the location also close to the middle of the jet, instead of the outside.

>>>>

A comprehensive description of the development of the propeller can be found in \cite{carlton2007}.  Here I won't elaborate on beyond describing some essential features and characteristics with which one ought to be familiar, primarily for the purpose of comparing different propellers and understanding their evolution into the systems such as pumpjets.

@@

\subsubsection{Pitch}
Perhaps one of the most important characteristics of a propeller is its pitch.  The pitch represents the distance that a blade section would travel forward if it carved its way around a helix through one full rotation.  It is intuitive to think of as something akin to the angle-of-attack of the propeller blade to the water, however this is technically misleading as the movement of the water-column incoming to the blade, as well as the speed of rotation, also have a significant bearing on what the actual angle of attack of the blade ends up being.  Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unitless equivalent, as is common in the description of many characteristics of propellers and propulsion systems.

@@

The selection of the exact blade selection that is used to define pitch must be specified as being at some fraction of the radius from the centre, given the tendancy for the blades to twist.  This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.  A thorough discussion of pitch can be found in \cite[35-37]{carlton2007}.

@@

Where $V_a$ is the speed of advance of the propeller, or the speed at which water arrives at the front-side of the propeller, and $n$ represents the rate of the propeller's rotation (revolutions per second) and $D$ is the propeller's diameter.  For fullness of understanding it is also worthwhile to note here that $V_a$ does not correspond directly to the vessel's speed of travel, though they are related. To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed \parencite[15]{MAN2017}.  Another key distinction for the consideration of jets later on is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field. In \cite[213]{lewis1988} $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.

One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$. This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced. This represents circumstances where a propeller encounters effectively zero load.  This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.  Since drag due to friction scales with velocity squared, very low speed do correspond to increasing advance ratios for a given propeller size, as would increases in propeller size for a given speed and pitch ratio. In practice however, under self-propuslion conditions for open propellers, slowing down towards does not elevate the advance ratio far enough past beyond peak efficiencies for overall efficiency to fall very far, as can be seen in Figure \ref{fig:EfficiencyBulten.png}

@@

\subsection{Ducted Propellers}
A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.  Ducted propellers fall into two categories, one of which is referred to frequently and interchangeably as a pump-jet.  The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating duct.  The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).

A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted. \footnote{It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.  But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear, and in practice ducts will still tend to decelerate the flow on approach to the impeller unless specifically shaped to narrow in the opening.}

@@

\subsubsection{The Accelerating Duct}
Accelerating ducts are most commonly used by vessels which have to operate at high loadings at very low speed, such as tugs, barges, or trawlers.  In those circumstances, relatively small vessels need to exert considerable forces against substantial loads, at very low water-speeds.  Also, in these circumstances, the practical size of a propeller might be constrained, since operating an optimally sized propeller might increase the complexity of machinery required to deliver the power to so low a hub in the water, and it might also be difficult to arrange for the thrust to be directed substantially against the centre of drag of the vessel.

The effect of the nozzle in these circumstances actually increase the optimal efficiency of the propeller, when operating at very low water-speeds but high thrusts (low advance ratio), since the pressure reduction on the inside of the nozzle induced by the movement of the water through the nozzle actually produces a net thrust forward, since the interior face at the opening of the nozzle is necessarily angled forward. This has the effect of distributing the suction force across a larger area, which now includes the interior of the nozzle.  An equivalent way of describing the same effect is that the nozzle causes the water to be accelerated more gradually, over a longer distance, by inducing it to move more quickly as it passes through to the entrance.  As a broad generalisation, more gradual or gentle actions tend to involve lower overall losses to turbulence, and hence the efficiency of the propeller can be increased.

A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.  Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.  It is also around this vortex that blade tip cavitation tends to occur. By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation, or the vortices that occur even when cavitation does not, can be diminished.

However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.  In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased.  In many circumstance cavitation can and does still occur in ducted propellers near the blade tips at design loads \parencite{moulijn2015}.

@@

A detailed explanation of the means by which the ideal efficiency of propeller can be improved can be found in \cite[213-222]{lewis1988}, as well as other texts on marine propulsion, such as \cite{carlton2007}. A substantial work exploring the efficiencies of a range of accelerating and decelerating ducts can also be found in \cite{oosterveld1970}.  A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low waterspeeds.  At higher waterspeeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.  When the exterior of the duct experiences relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.  Furthermore, in general these models do not incorporate the effect of flow separation, which is acknowledged as a possibility on both the exterior and interior of ducts, which can dramatically reduce efficiency performance (\cite[20]{oosterveld1970}, \cite[214]{lewis1988}).

@@

In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.

The addition of some new parts such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.  Not all decelerating ducts have stators, but the impact of the duct on the stream will remain. The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.  In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to lessen vorexs and cavitation on the extremities.  However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.  In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.  No definitive transition point seems to be accepted as having great significance.  Most commonly literature will refer to a 'rotor', a more general term which could encompass both propellers and impellers.

\subsubsection{Waterjets}
The use of the word waterjet, as opposed to pumpjet, has a tendency signify that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.  The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.  However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump. Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).  There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as 'submerged waterjets' \parencite{buckingham2008submarine}.

@@

\subsection{Recent Commentary}
There has been already some attention given to the probable efficiencies of pump jets at low speed. Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.  This argument relies on pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.  As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.  This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pump-jet for a submarine might perform in quantitative terms, relative to a propeller.

There has also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017}, and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.  Whilst the statement that the state of the art is much advanced is obviously true in the case with marine propulsion as with most sciences, it is also equally true that new advances tend to confirm, rather than annul, many of the works that have been previously undertaken.  To the extent that modern literature still supports the overall constraints and bounds provided by some of the foundational earlier works, a great deal of useful information can still be obtained from earlier works.  In particular, the foundational theories, often derived from well-established principals of conservation of momentum, energy, as applied by the likes of Bernoulli and Euler, are as well respected and deeply depended upon today as they were centuries earlier when they were first arrived at.  It is noteworthy that some of oldest works, such as those by Wislicenus and Henderson, are still amongst the most frequently cited in the most recent of papers.

\subsection{Modern literature and results}
Abundant modern works, including those which involve experimental validation of theoretical results also clearly confirm that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.  It should be noted however, that relatively little literature focuses on the performance of propulsive systems far far outside of their intended design speeds, which tends to be far higher than the patrol and transit speeds of diesel electric submarines.  Fewer still tend to express their work in terms of ship-speed, instead adopting the convetion of using dimensionless ratios so that work done on one system can be readily translated into different combinations loads, pressures, speeds and sizes.  However, from a few notable studies exist which provide important reference points. (Further interpretation of the what we can lean from the trends expressed in dimensionless coefficients will be discussed later.)

@@

Both of these previous studies confirm the general and widespread assumption that jets are known to be more efficient at high speeds, and propellers at lower speeds.  They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.  However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets) do have at least one significant difference in their makeup, though their internal pump mechanisms may often be similar. Specifically, pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.  Given that these factors might affect efficiency, and the overall design of the nozzle considerably, more direct comparison with fully submerged jets should be sought to make more direct comparisons.

@@

The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.  The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.  In the case of this analysis, modifications to the hullform in order to accommodate the jet (which are likely to be more substantial for a surface ship than a submarine) also have an impcat on the resistance offerd by the hull.  This also has an interaction with the different shape of the propulsors, which alters the required level of delivered (effective) power to achieve a given speed, in addition to the considerations of how efficiently the propulsor generates that effective power. The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form. If the power could be reduced by 5\% (e.g. by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hullform needed to accommodate the jet, rather than the jet propulsion efficiency itself \parencite{giles2010}.  As the propulsive efficiency curves given in the study show, the propulsors themselves are considerably further from reaching parity, still separated by about 10\% at 25kt.

The paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.  At 5kt, the propeller was more than twice as efficient as the waterjet.  This study has been used to directly inform the 'central' efficiency curves assessed in this analysis.

@@

It appears from recent examples in the literature that there is no evidence of jets approaching the same efficiency as propellers at speeds as low as 10kt.  Indeed, there seems to be considerable evidence that jets of most kinds suffer substantial efficiency penalties relative to propellers in this regime, and that this is understood and accepted as an inevitable design trade-off to be made.  In the case of a submarine, which might spend nearly 100\% of its time travelling at 10kt or less, this disadvantage (as opposed to less than 35\%, as assumed for a surface ship in \cite{giles2010}) would become extremely substantial, which will be quantified in later sections.

Moreover, recent literature seems to confirm that efficiency of pumpjets fall off precipitously towards zero the closer that the vessel speed comes to zero.  This strongly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.  To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.  This should further limit the range of error that might be made in estimating realistic bounds of performance for a system which we do not have precise design specifications for.

@@

Pumpjets are necessarily low in efficiency at very low speeds because they attempt to get the water to do multiple things.  They don't just accelerate a water colum.  They first slow it down (increasing its pressure), then add pressure to it, then accelerate it.  When the total amount of energy embodied in the incoming water stream, and the total amount of energy you want to add to the water both get very small (as they do at very low speed) there simply isn't enough energy to do all things at once to all of the water.  The energy penalty inflicted by using the additional devices becomes very large in proportion, and part of the water doesn't do all the necessary things for the jet to work.  In essence, if you try to slow down something that's already going slowly, it tends to stop  That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, begins to take place.  When there isn't enough energy to move all the water through the pressure-hurdles imposed by the duct, only some of it moves, and the rest spins in place.

There are other some fundamental limitations on the peak efficiency that can be achieved by jets based on simple calculations of ideal efficiency from momentum theory, which are discussed in the next subsection.  They also provide enduring, reliable limitations on the extent to which ingenuity or design changes might be able to improve the efficiency of jets at low speed. But the primary reason why jet efficiencies decline towards zero at low speed is to do with flow separation \parencite{fujisawa1995}, which is what occurs when the assumptions inherent in momentum theory's analysis of an 'ideal' actuator disc start to break down.

The conditions which lead to flow separation in a pumpjet can be relatively simply explained in terms of conservation of energy, and conservation of mass. Conservation of mass requires that as a pipe or duct expands, the flow through it necessarily slows down, in order to occupy the large volume, in proportion to the new cross-sectional area.  Conservation of energy requires that it increases in pressure, and defines a very specific relationship between the pressure increase and the change in velocity, which is determined by the shape of the duct.  That relationship says that the static pressure increase is inversely proportional change in velocity \textbf{squared}, as given by Bernoulli's equation, which was discussed in earlier sections.  As a result, at high speeds, the total energy of the system is dominated by its velocity, but at very low speeds the static pressure is much more significant.

It also means that for any given change in velocity, the static pressure changes also, but doesn't change as much as the velocity change.  The consequences for this pumpjets are and inevitable, since the geometry of the duct is inevitably fixed.  At very high flow rates, the expansion of the area ahead of the impeller creates an adverse pressure gradient, as the water must slow down to fill the greater area.  This adverse pressure gradient is larger at higher speeds, the proportions aren't linear, hence the the speed is much much larger, due to the square relationship, and flow is consistent (though still turbulent) throughout the entire cross-sectional area of the duct.

@@

The same phenomenon is also prone to occur near roots of the blades of the rotor \parencite[15]{henderson1964}.  Necessarily, the blades of the rotor are doing work on the fluid, so unless the impeller is in a region of considerable contraction, pressure will necessarily have to increase significantly through the blades, providing an adverse pressure gradient, which again must be constant radially.  Since closer to the hub the velocity of the blades through the water is lower, the blades tend often to also be longer and deflect the flow even more in order to do the same amount of work on the fluid. With the longer surface over which the fluid must travel, as well as greater angle of deflection, flow separation is prone to occur here too.  The physics will be exactly analogous to that of the separation on the inside of the duct intake, except the pressure gradient will be provided by a different source (the rotor doing work), and the surface (the blade roots) will also be different. This will also mean that the direction of rotation of the vortices will be different, and the location also close to the middle of the jet, instead of the outside.
